Continuous gravitational waves in the lab: recovering audio signals with a table-top optical microphone

### 2020-09-11 paper/ifo-singleTone.tex
Deeksha's comments: single and wandering tone

\begin{figure}
	\includegraphics[width=.49\textwidth]{figures/webcam_expt_4_0209-cropped.pdf}
	\caption{\label{fig:webcam_spectrum}
Injected tone at a constant frequency: Fourier amplitude of the intensity time series recorded by the webcam. 
The frequency of the injection is $2.09\,{\rm Hz}$. 
The recovered signal peaks at $2.099\,{\rm Hz}$ with a FWHM of $0.033\,{\rm Hz}$. 
There are also two visible harmonic peaks at integer multiples of the main peak. 
Their amplitudes as a fraction of the height of the main peak are  $0.103$ and $0.098$ for the $4.19\,{\rm Hz}$ and $6.28\,{\rm Hz}$ peaks respectively. 
}

\end{figure}

Continuous-wave searches look for signals which are close to monochromatic. 
In this section we consider a simple sinusoidal tone; a note at a single constant frequency. 
As described in Section~\ref{sec:ifo}, the audio is played through a speaker fixed to the back of one of the mirrors.

The intensity of the interference pattern is measured at a single point on the screen using a commercial USB webcam placed to view the screen (see Fig.~\ref{fig:ifo_schematic_webcam}). 
The webcam's sample rate is $30\,{\rm Hz}$, which limits the spectral content of observable signals to below $15\,{\rm Hz}$, the Nyquist frequency.
A tone under $15\,{\rm Hz}$ is played through the speaker for one minute and the interference pattern recorded with the webcam.

We take data from a single off-centre pixel indicated by the red dot in Fig.~\ref{fig:interference_pattern}. 
The green-channel of the video is used as an approximation to the total intensity as the laser produces monochromatic green light (at $532\,{\rm nm}$).
As an initial inspection, we play the timeseries of the intensity using the \textbf{io.wavfile.write} function from the Scipy~\cite{scipy} package in Python~\cite{python}. 
The two tones sound similar by ear and don't produce any noticeable beating. 
The result is quantified in Fig.~\ref{fig:webcam_spectrum}. 
A tone of $2.09\,{\rm Hz}$ is played through the speaker and the data recorded via webcam. 
Figure~\ref{fig:webcam_spectrum} shows the Fourier transform of the recording.
We measure a peak amplitude at $2.099\,{\rm Hz}$ with a full width half maximum (FWHM) of $0.033\,{\rm Hz}$. 
Two harmonics can also be seen in Fig.~\ref{fig:webcam_spectrum} at integer harmonics of the main peak. 
The $4.19\,{\rm Hz}$ and $6.28\,{\rm Hz}$ peaks have amplitudes of $0.103$ and $0.098$ respectively as a fraction of the height of the main peak. 
They are probably due to the nonlinear system response discussed in Section~\ref{sec:ifo} and Appendix~\ref{app:intensity_derivation}.

>>>>

\begin{figure}
	\includegraphics[width=.49\textwidth]{figures/webcam_expt_4_0209-cropped.pdf}
	\caption{\label{fig:webcam_spectrum}
Recovery of a tone at a constant frequency: Here we plot the Fourier amplitude of the intensity pattern against frequency.
The injected signal has a frequency of $2.09\,{\rm Hz}$, while the recovered signal peaks at $2.099\,{\rm Hz}$ and has a FWHM of $0.033\,{\rm Hz}$.
The plot also shows two harmonics at $4.19\,{\rm Hz}$ and $6.28\,{\rm Hz}$ with normalised amplitude to the primary peak of $0.103$ and $0.098$ respectively.
}	
\end{figure}

Continuous-wave searches look for nearly monochromatic signals. In this section we consider a simple sinusoidal tone; a signal at a single constant frequency.

As described in Section~\ref{sec:ifo}, the audio signal is played through a speaker fixed to the back of M2. The intensity of the interference pattern is measured at a single point on the screen using a commercial USB webcam (see Fig.~\ref{fig:ifo_schematic_webcam}). The webcam samples at a rate of $30\,{\rm Hz}$, which limits the spectral content of observable signals to below $15\,{\rm Hz}$, the Nyquist frequency. A tone with a frequency of 2.09Hz is played through the speaker for one minute and the interference pattern is recorded by the webcam.

We take data from an off-centre pixel as indicated by the pink star in Fig.~\ref{fig:interference_pattern}. The pixel intensity in the green-channel of the video (the webcam records in three colour channels: red, blue, and green) is used to approximate the total intensity of the interference pattern. This is because the laser emits monochromatic green light at $532\,{\rm nm}$.
The amplitude of the Fourier transform of the recorded signal is shown in Fig.~\ref{fig:webcam_spectrum}. We measured a strong signal at $2.099\,{\rm Hz}$ with a full width half maximum (FWHM) of $0.033\,{\rm Hz}$. Although this range doesn't contain our injected frequency, the error is not substantial and cannot be heard by the (untrained) human ear.
Two harmonics can also be seen at integer multiples of the peak frequency. The peaks at $4.19\,{\rm Hz}$ and $6.28\,{\rm Hz}$ have amplitudes of $0.103$ and $0.098$ as a fraction of the height of the main peak, respectively. They are likely due to the nonlinear response of the system discussed in Section~\ref{sec:ifo} and Appendix~\ref{app:intensity_derivation}.

In the following section, we use a hidden Markov model technique, commonly used in continuous wave searches, to recover a slowly wandering frequency signal.

### 2020-09-11 paper/ifo-tableTopGWs.tex
Deeksha's comments: single and wandering tone

The basic physics of gravitational-wave detection can be demonstrated with a table-top Michelson interferometer, an optical configuration commonly used in undergraduate labs. 
A simple Michelson interferometer comprises a laser, a beamsplitter, and two mirrors. 
Laser light is split by a beamsplitter into two arms at right angles. 
At the end of each arm is a mirror which reflects the beams back to the beamsplitter where they recombine. 
The combination of the two beams produces an interference pattern which is dependent on the relative distance travelled by each of the beams. 
A change in arm length results in a change in the interference pattern.

Gravitational-wave detectors are large, complex interferometers.
The LIGO and Virgo detectors have $4\,{\rm km}$ and $3\,{\rm km}$ arms respectively.
They are sensitive to minuscule changes in distance, for example the first detection of a binary black hole merger had a peak gravitational-wave strain of $10^{-21}$ ~\cite{GW150914}.
However, their design is based fundamentally on the Michelson interferometer. 
Detection and analysis concepts can be demonstrated with table-top optical equipment. 
See Refs.~\cite{TTExhibit:2020,TTExhibit:online,AMIGO:online,ThorLabsIFO,LIGOIFOGlue,LIGOIFOMagnets,FoxEtAl:1999} for a selection of table-top Michelson interferometer designs used for science communication.

A commonly-used analogy in explaining gravitational-wave science is sound. 
Signals of binary black hole and neutron star mergers can be converted into audio as an aid to explanations~\cite{SoundsOfSpaceTime:online,BlackHoleHunter:online}~\footnote{When introducing the acoustic analogy to lay audiences, it is important to emphasize that gravitational waves are not sound. For example, gravitational waves can propagate in a vacuum, whereas sound cannot.}.
Sound vibrations provide a simple means to move the components of a demonstration interferometer. 
The vibrations move the mirrors, changing the length of the interferometer arms, and the resulting interference pattern. 
We use sound throughout this work as a means of simulating gravitational-wave-like signals and the data analysis and signal processing techniques used to detect them.

\begin{figure}
	\includegraphics[width=0.49\textwidth]{figures/ifo_schematic_webcam_edit.pdf}
	\caption{\label{fig:ifo_schematic_webcam}
Schematic of the table-top Michelson interferometer. 
The laser (top-left) is a Class 2 $532\,{\rm nm}$ laser~\cite{ThorLabsIFO}. 
The beam is first reflected from a mirror at a right angle and then focused at a converging lens before reaching the beamsplitter which is $50$/$50$ in reflection and transmission. 
The split beams are reflected from mirrors (M1 and M2) at the ends of the interferometer arms and are recombined at the beamsplitter. 
The beam is enlarged by a diverging lens after the beamsplitter. 
The interference pattern is projected onto a screen and recorded using a commercial webcam connected to a personal computer (PC). 
A speaker fixed to the back of M2 provides the signal input from the PC.  
    }
\end{figure}

The Michelson interferometer configuration used in this work is shown schematically in Fig.~\ref{fig:ifo_schematic_webcam}.
It is constructed on a $450 \times 450 \,{\rm mm} $ optical breadboard. 
The laser (top-left) has a wavelength of $532\,{\rm nm}$.
In the configuration shown in Fig.~\ref{fig:ifo_schematic_webcam} the laser beam is first reflected from a mirror which turns the beam $90^{\circ}$ (to save space on the optical breadboard). 
The beam passes through a converging lens and then to the beamsplitter, which is $50$/$50$ reflection and transmission. 
The beams are reflected by mirrors at the end of each of the two arms ($\approx 7.5\,{\rm cm}$ and $\approx 10.0\,{\rm cm}$ long) and are recombined at the beamsplitter. 
The resulting interference pattern is projected onto a screen and recorded by a webcam.

\begin{figure}
 \begin{center}
  \includegraphics[width=.45\textwidth]{figures/webcam_still0_crop.pdf}
 \end{center}
 \caption{\label{fig:interference_pattern}
Michelson interference pattern as viewed by an off-axis webcam (which is why the pattern is elliptical). 
The red dot marks where the intensity time series is measured. 
The central bright fringe of the interference pattern is approximately $5\,{\rm mm}$ in diameter. 
}
\end{figure}

The interference pattern is a ring of interference fringes as shown by the photograph in Fig.~\ref{fig:interference_pattern}. 
The pattern is enlarged by a diverging lens between the beamsplitter and the screen to aid in viewing. 
A change in the relative length of the interferometer arms results in the interference rings moving radially inwards or outwards. 
Ring interference patterns like this are often used in gravitational-wave table-top demonstrations to allow changes in the pattern to be easily viewed.

A $0.5\,{\rm W}$ speaker is fixed to the back of one of the mirrors (M2 in Fig.~\ref{fig:ifo_schematic_webcam}), providing a controllable source of vibrations. 
The speaker is one half of a pair of commercial, USB-powered speakers fed by a $3.5\,{\rm mm}$ jack and driven by a PC. 
It is attached to the back of the mirror using commercial adhesive putty. 
The other speaker in the pair is kept face-down and as far away from the interferometer as the cables allow to reduce its impact as a second source of vibrations.

The motion of the interference fringes follows the motion of the mirror, and therefore the speaker. 
If the fringe motion is small, then changes in the intensity at some point on the screen occur at the same frequency as the injected audio but with amplitude given by a transfer function that accounts for the coupling and resonance of the speaker-mirror system.
If the fringe motion is large enough for multiple fringes to pass over the measured point in a single speaker deformation, then over-counting artificially raises the measured frequency. As such, any motion of the fringes must be kept small by playing sound softly through the speaker.
Even without over-counting, large fringe motions introduce a nonlinear relation between the intensity fluctuations and injected audio, leading to troublesome---but physically interesting!---phenomena like frequency doubling.
We derive the nonlinear relation between the separation of the mirrors and the intensity of the pattern at some point on the screen in Appendix~\ref{app:intensity_derivation}.

Data can be recorded via a webcam (as shown in Fig.~\ref{fig:ifo_schematic_webcam}) or photodiode. 
In Sections~\ref{sec:single_tone} and~\ref{sec:viterbi_wandering} a webcam is used. 
In Section~\ref{sec:optical_microphone}, which considers more complex audio, we turn to a photodiode because it offers a higher sample rate than the webcam.

>>>>

Gravitational-wave detectors such as aLIGO and aVirgo are large, complex experiments. However, their design is based fundamentally on a Michelson interferometer, an optical configuration commonly used in undergraduate laboratories.
It consists of a beam splitter which splits light from a laser down two long perpendicular ``arms'' ($4\,{\rm km}$ long at aLIGO, $3\,{\rm km}$ at aVirgo).
Mirrors at the end of each arm reflect the light back towards the beam splitter, where they recombine and produce an interference pattern.
This pattern is dependent on the relative distance travelled by each beam and thus on the relative length of each arm. The current generation gravitational-wave observatories use this technique, in combination with others, to detect minuscule changes in distance of two arms due to a passing gravitational wave.
For example, the first detection of a binary black hole merger changed the relative length of the two arms by $10^{-21}$~\cite{GW150914}, smaller with the width of a proton.

The complex detection and analysis techniques used in the gravitational wave community can be demonstrated using a table-top experiment, such as the one presented here.
Refs.~\cite{TTExhibit:2020,TTExhibit:online,AMIGO:online,ThorLabsIFO,LIGOIFOGlue,LIGOIFOMagnets,FoxEtAl:1999} provide a selection of other table-top Michelson interferometer designs used for science communication. 
In addition to this, gravitational-wave signals from binary black hole and neutron star mergers are often converted to audio signals~\footnote{When introducing the acoustic analogy to lay audiences, it is important to emphasize that gravitational waves are not sound. For example, gravitational waves can propagate in a vacuum, whereas sound cannot.} to aid in explanations.~\cite{SoundsOfSpaceTime:online,BlackHoleHunter:online}.
This analogue allows scientists to use sound vibrations as a simple means to move the components of a demonstration interferometer. The vibrations move the mirrors, changing the length of the interferometer arms, and thus changing in the interference pattern.
We use audio signals throughout this work to simulate gravitational-wave-like signals and to demonstrate different data analysis and signal processing techniques used to recover such signals.

\begin{figure}
	\includegraphics[width=0.49\textwidth]{figures/ifo_schematic_webcam_edit.pdf}
	\caption{\label{fig:ifo_schematic_webcam}
Schematic of the table-top Michelson interferometer. 
The output of a Class 2 $532\,{\rm nm}$ laser~\cite{ThorLabsIFO} (top-left) is reflected from a mirror at a right angle. The beam then passes through a converging lens before reaching the beamsplitter which is $50$/$50$ in reflection and transmission. 
The split beams are reflected from mirrors (M1 and M2) at the ends of the interferometer arms and are recombined at the beamsplitter. 
The output interference pattern is enlarged using a diverging lens, projected onto a screen, and recorded using a commercial webcam connected to a personal computer (PC). 
A speaker fixed to the back of M2 is used to inject audio signals from the PC into the interferometer.
    }
\end{figure}

The optical configuration of the Michelson interferometer used in this work is shown in Fig.~\ref{fig:ifo_schematic_webcam}.
It is assembled on a $450 \times 450 \,{\rm mm} $ optical breadboard and uses as a source a green eye-safe laser (top-left) with peak emission at $532\,{\rm nm}$.
The output of the laser is first reflected off a mirror which turns the beam $90^{\circ}$ (to save space on the optical breadboard and to have greater control of the alignment of the interferometer).
After passing through a converging lens of focal length (FL = \jam{???}), the laser beam is incident on a beamsplitter, which reflects half of the light towards mirror 1 (M1) and transmits the other half to mirror 2 (M2). 
A $0.5\,{\rm W}$ speaker is fixed to the back of M2 using commercial adhesive putty and serves as a controllable source of vibrations. This speaker is one of a pair of commercial, USB-powered speakers, fed by a $3.5\,{\rm mm}$ jack and driven by a PC. The other speaker in the pair is kept face-down and as far away from the interferometer as possible to prevent it from acting as a second source.
The beams are reflected by mirrors M1 and M2, located at the end of (approximately) $7.5\,{\rm cm}$ and $10.0\,{\rm cm}$ long arms.
The beams recombine at the beamsplitter and produce an interference pattern that is then enlarged using a diverging lens of FL = \jam{???}, and projected onto a screen.
The interference pattern, as shown by the photograph in Fig.~\ref{fig:interference_pattern}, is a set of concentric light and dark rings (fringes). 
A change in the relative length of the arms causes these rings to move radially in or outwards.
% Ring interference patterns like this are often used in gravitational-wave table-top demonstrations to allow changes in the pattern to be easily viewed. 
An intensity time series of some point in the interference pattern is recorded by a webcam (or later, a photodiode) for data analysis to extract the signal injected at the speaker.

\begin{figure}
 \begin{center}
  \includegraphics[height=.45\textwidth, angle=-90]{figures/webcam_still0_star.pdf}
 \end{center}
 \caption{\label{fig:interference_pattern}
The interference pattern produced by the table-top Michelson interferometer.
This image was taken with the webcam off the beam axis, which may explain why the (predicted spherical) pattern appears elliptical.
The pink star indicates the point where the intensity time series measurements were taken.
The central bright fringe of the interference pattern is approximately $5\,{\rm mm}$ in diameter. 
}
\end{figure}

The motion of the interference fringes follows the motion of M2, and therefore of the speaker.
The amplitude of these motions is given by a transfer function that accounts for the coupling and resonance of the speaker-mirror system.
If the relative change in the length of the arms is kept small enough, the intensity of the interference pattern at any point on the screen oscillates at the same frequency as the injected audio.
But if the relative change in the length of the arms is large enough for multiple fringes to pass over the measured point in a single speaker deformation, then over-counting (artificially) raises the measured frequency.
As such, any motion of the fringes must be kept small by playing the sound softly through the speaker.
Even without over-counting, large fringe motions display a nonlinear relation between the intensity fluctuations and injected audio, leading to troublesome---but physically interesting---phenomena (e.g frequency doubling).
We derive this nonlinear relationship between the length difference of the arms and the intensity of the pattern at some point on the screen in Appendix~\ref{app:intensity_derivation}.

In the following section, we start simply by injecting and recovering a audio signal of constant frequency (a single tone).

% Data can be recorded via a webcam (as shown in Fig.~\ref{fig:ifo_schematic_webcam}) or photodiode. 
% In Sections~\ref{sec:single_tone} and~\ref{sec:viterbi_wandering} a webcam is used. 
% In Section~\ref{sec:optical_microphone}, which considers more complex audio, we turn to a photodiode because it offers a higher sample rate than the webcam.

### 2020-09-11 paper/ifo-wanderingTone.tex
Deeksha's comments: single and wandering tone

Continuous gravitational wave signals may wander slowly in frequency over time, due to stochastic internal processes in the superfluid interior of isolated neutron stars~\cite{MelatosDouglassSimula:2015,Jones:2010} or stochastic accretion flows in binary neutron stars~\cite{BildstenTB:1998}. 
The audio analogue of this is a tone that wanders in frequency; a note that changes pitch.

The analysis techniques used here are inspired by those used to search for continuous waves from spinning neutron in LIGO and Virgo data~\cite{SuvorovaEtAl:2016,SuvorovaEtAl:2017}.
Continuous wave searches are performed on long datasets (months to years in duration). 
The frequency of gravitational-wave emission wanders significantly over the observation period. 
In this context, ``significantly'' means ``across multiple frequency bins'', where the typical width of a frequency bin is the reciprocal of the total observing time~\cite{JKS:1998,ScoX1O2Viterbi:2019}

One method to search for a wandering signal is to split the data into several shorter segments which are analysed individually. 
The detection statistics in every segment are combined to form a grid in time and frequency (spectrogram). 
The value at each spectrogram point determines the likelihood that a signal is present at a particular time and frequency. 
In gravitational-wave data analysis, a detection statistic is used to calculate the likelihood of a signal given the antenna beam pattern of the detector, which varies as the Earth rotates and orbits the Sun~\cite{JKS:1998}.
The detection statistic used depends on the type of target~\cite{JKS:1998,SuvorovaEtAl:2017}. 
In this work the detection statistic is the amplitude (modulus) of the Fourier transform of the data at each time segment, which maximizes the likelihood of detecting a sinusoid in Gaussian noise, as described in Appendix~\ref{app:sinusoid_likelihood}. 
The resulting spectrogram gives the Fourier amplitude for each frequency and time bin.

In continuous wave searches a hidden Markov model is used to describe the system. 
The frequency (the state) of the signal is unknown (hidden) and the signal is Markovian in that the state at any time depends only on the state in had at the previous time step. 
The hidden Markov model method relates the observable quantity (the detection statistic) to the signal frequency (the hidden state). 
The task is to find the most likely wandering frequency signal given the observable of the spectrogram. 
The Viterbi algorithm~\cite{Viterbi:1967} is an efficient method to find the most likely path of a signal given an observation sequence. 
This method is used in continuous wave searches for a range of astrophysical targets~\cite{ScoX1O2Viterbi:2019,ScoX1ViterbiO1:2017,MillhouseStrangMelatos:2020,JonesSun:2020,MiddletonEtAlO2LMXBs:2020,PostMergerRemnantSearch:2019,SunEtAlSNR:2018,viterbi_application}.

>>>>

Continuous wave searches, such as in Refs.~\cite{SuvorovaEtAl:2016,SuvorovaEtAl:2017,SearchTwoSpecS6:2017,SunEtAlSNR:2018,JonesSun:2020}, are often (not always) performed on long datasets (months to years in duration).
The frequency of a continuous wave signals may wander significantly over this timescale, due to stochastic internal processes in the superfluid interior of isolated neutron stars~\cite{MelatosDouglassSimula:2015,Jones:2010} or variable accretion flows onto a neutron star from a stellar companion (e.g. as in LMXBs)~\cite{BildstenTB:1998}.
In this context, wandering ``significantly'' means ``across multiple frequency bins'', where the typical width of a frequency bin is the reciprocal of the total observing time~\cite{JKS:1998,ScoX1O2Viterbi:2019}.
The audio analogue to these continuous wave signals is a tone that wanders in frequency; a note that changes pitch.
Here, we adapt the data analysis technique used to search for slowly wandering continuous waves signals from spinning neutron stars in aLIGO and aVirgo data~\cite{SuvorovaEtAl:2016,SuvorovaEtAl:2017}.

One method to search for a wandering signal is to split the time series into several shorter segments, analyse these individually, then recombine them (under some continuity assumption) to coherently analyse the whole.
The detection statistics in every segment are combined to form a grid in time and frequency (spectrogram).
The value at each grid point gives the likelihood that a signal is present at that particular time and frequency.
In gravitational-wave data analysis, a detection statistic is used to calculate the likelihood of a signal given the antenna beam pattern of the detector, which varies as the Earth rotates and orbits the Sun~\cite{JKS:1998}.
The detection statistic used depends on the type of target~\cite{JKS:1998,SuvorovaEtAl:2017}.

In this work the detection statistic is the amplitude (modulus) of the Fourier transform of the data at each time segment, which maximizes the likelihood of detecting a sinusoid in Gaussian noise, as described in Appendix~\ref{app:sinusoid_likelihood}.
The resulting spectrogram gives the Fourier amplitude for each frequency and time bin.
In continuous wave searches a hidden Markov model is used to describe the system.
The frequency (the state) of the signal is unknown (hidden) and the signal is Markovian in that the state at any time depends only on the state at the previous time step.
The hidden Markov model method relates the observable quantity (the detection statistic) to the signal frequency (the hidden state).
The task is to find the most likely wandering frequency signal given the observable of the spectrogram.
The Viterbi algorithm~\cite{Viterbi:1967} is an efficient method to find the most likely path of a signal given an observation sequence.

This method is used in continuous wave searches for a range of astrophysical targets~\cite{ScoX1O2Viterbi:2019,ScoX1ViterbiO1:2017,MillhouseStrangMelatos:2020,JonesSun:2020,MiddletonEtAlO2LMXBs:2020,PostMergerRemnantSearch:2019,SunEtAlSNR:2018,viterbi_application}.
